The Geometry of Archaeological Thought: Analysing discplinary change using word embeddings from conference abstracts

### 2020-05-08 documents/paper.Rmd
exporing word prop changes by two periods

---
title: "The Geometry of Archaeological Thought: Analysing discplinary change using word embeddings from conference abstracts"
output: bookdown::html_document2
bibliography: bibliography.bib
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

@@

In this paper we present a dataset of all available abstracts of papers and posters presented at annual meetings of the Society of American Archaeology. We demonstrate the use of word embeddings, a commonly used computational tool in natural language processing (NLP) and machine learning, as a method to measure, quantify, and compare changes in the way archaeologists write. We present the temporal reference plot as a method for visualizing disciplinary change in archaeology. As a case study, we use these data and methods to explore how archaeologists' writing about explaining and theorizing the past have changed over time.

@@

- source of the data
We collected abstracts from Society of American Archaeology (SAA) annual meeting programs from 1964 to 2020. The total number of abstracts are 68,176 from 40 annual meetings excluding some years without available abstracts. The source of abstract includes papers, posters, workshop posters, symposium represented by paragraphs. The annual programs are available online at https://www.saa.org/annual-meeting/annual-meeting-archives/program-archives. The original format of abstracts on websites includes scanned images for early years and accessible PDF after 2004.

- preparation steps: data cleaning (Cite the quanteda pkg, and things they cite)

- word embeddings (cite our readings)
- temporal referencing plots (cite our readings)

>>>>

---
title: "The Geometry of Archaeological Thought: Analysing discplinary change using word embeddings from conference abstracts"
output: bookdown::html_document2
bibliography: references.bib
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

@@

In this paper we present a dataset of all available abstracts of papers and posters presented at annual meetings of the Society of American Archaeology. We demonstrate the use of word embeddings, a commonly used computational tool in natural language processing (NLP) and machine learning, as a method to measure, quantify, and compare changes in the way archaeologists write cf. [@Garg2018; @Kozlowski2019]. We present the temporal reference plot as a method for visualizing disciplinary change in archaeology. As a case study, we use these data and methods to explore how archaeologists' writing about explaining and theorizing the past have changed over time.

@@

- source of the data

The data for our study comes from the text of oral and poster presentation abstracts presented at the annual meetings of the Society of American Archaeology (SAA). The earliest meeting in our data is 19XX, and we have some years missing, especially XXXX-XXXX. The total number of abstracts is 68,176 (xxx words) from 40 annual meetings. We obtained PDF files of annual programs from https://www.saa.org/annual-meeting/annual-meeting-archives/program-archives.

The PDFs for the years XXXX-XXXX contain images of pages, and after 2005 the PDFs are searchable text. We converted the page images from PDF to PNG files using ImageMagick software, and then used optical character recognition to extract text from the PNG files with the Tesseract package [@Kay2007]. We obtained an acceptable rate of 95-98% word-level OCR accuracy across the PDFs [citation]. After OCR, we used the quanteda package  [@Benoit2018] to remove punctuation, numbers, and stopwords, such as "the", "an", "a", [specifically why do we remove stopwords?].

- something about basic steps
-- transform text into dfm, define dfm...

- word embeddings (cite our readings)
-- a few sentences to introduce the reader to how WE have been used in other recent studies. this is a little bit background. '...recently used in many social science fields to...' tell the reader what was the finding of some key studies using WE that have inspired us.

<!-- There are two major word-level pretrained word embeddings, Word2Vec and Glove according to different methods used for training [@Rong2014; @Pennington2014]. -->

-- then some more detail about the mechanics of the method, as we are using it. Say how what we did works in general.

In this article, we trained word embeddings with the GloVe model using th text2vec package [@Selivanov2016]. The GloVe model calculates ... creates the co-occurrence matrix.

- temporal referencing plots (cite our readings)
